A boundary-integral framework to simulate \\ viscous erosion of a porous medium

### 2018-04-11 docs/methods/erosion_methods.tex
small changes to language and formatting

Recent work in the high-Reynolds-number regime has highlighted the importance of a shape-flow feedback that occurs during erosion~\cite{ris-moo-chi-she-zha2012, moo-ris-chi-zha-she2013, hewett2017evolution, moore2017riemann} and the related processes of dissolution and melting~\cite{Huang2015, kondratiuk2015steadily, rycroft2016asymmetric, cohen2016erosion, hewett2017pear, claudin2017dissolution}.  In these cases, the flow alters the morphology of the immersed structures, which in turn modifies the surrounding flow, and so on. Such processes of `sculpting with flow' \cite{ristroph2018sculpting} have been examined in the context of a single body \cite{ris-moo-chi-she-zha2012, moore2017riemann} and, more recently, multiple bodies whose shape evolution is coupled through the flow \cite{hewett2017evolution}.
From this viewpoint, the example of an eroding porous medium presents new opportunities and challenges. First, since the medium is composed of many individual constituents, multi-body interactions must be considered. Second, the long-range effects inherent to Stokes flow implies that the shape-evolution of bodies are even more strongly coupled than in any other regime.
% in particular close-contacts?

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\paragraph{Contributions}
To perform this simulation of Stokesian erosion of a porous media, two fast and accurate methods are employed---one to solve the fluid equations and the other to integrate the eroding grains.  The fluid equations are reformulated in terms of a boundary integral equation (BIE)~\cite{poz1992}.  Since the boundary conditions are Dirichlet, we use a double-layer potential formulation so that the resulting BIE is second-kind.  This second-kind BIE is discretized with the spectrally accurate trapezoid rule~\cite{tre-wei2014} which results in a well-conditioned dense linear system for an unknown density function, and the linear system is solved using fast-multipole-accelerated GMRES~\cite{saa-sch1986}.  Once the density function is computed, the shear stress is computed with a spectrally accurate odd-even quadrature rule~\cite{sid-isr1988}.  In addition, the pressure and vorticity are computed in the fluid bulk (Fig.~\ref{fig:50bodies}) with an upsampled trapezoid rule to assist with visualizing the effect of erosion in porous media.

@@

\paragraph{Related Work}
In closely related work, Mitchell \& Spagnolie used a traction integral equation formulation to simulate viscous erosion of a single body in three dimensions \cite{mit-spa2016}. The flexibility of their formulation enabled simulations of a {\em mobile} body eroding in an {\em aribtrary} background flow (e.g.~uniform flow, shear flow, etc.). Another advantage of the traction formulation is that is directly provides the shear stress that is needed to evolve surfaces.  In the present work, we aim to simulate the erosion of a porous medium and, further, choose to fix the bodies to their initial locations. Thus, rather than a {\em single, mobile} body as considered in \cite{mit-spa2016}, we aim to simulate the erosion of {\em many, immobile} bodies. Although the traction formulation has several attractive features, we instead employ a standard double-layer formulation since it allows us to easily calculate bulk quantities (e.g.~vorticity, pressure, and velocity) that will assist our understanding of the erosion process.

@@

Since the double-layer potential is not able to represent force- and torque-free velocity fields,  we complete the double-layer potential with Stokeslets and rotlets~\cite{pow-mir1987, pow1993}
\begin{align}
  S[\llambda_\ell,\cc_\ell] = \frac{1}{4\pi} \left(-\log \rho \II + 
    \frac{\rr \otimes \rr}{\rho^2} \right)\llambda_\ell
  \quad \text{and} \quad
  R[\xi_\ell,\cc_\ell] = \frac{\rr^\perp}{\rho^2}\xi_\ell\, , 
  \quad \ell=1,\ldots,M \, ,
  \label{eqn:stokeslet_rotlet}
\end{align}
where  $\rr = \xx - \cc_\ell$, $\cc_\ell$ is a point inside the $\ell^{th}$ grain, $\rr^\perp = (r_2,-r_1)$, and $\rho = \|\rr\|$.  Then, the solution of~\eqref{eqn:stokes} is
\begin{align}
  \label{eqn:completed_DLP}
  \uu(\xx) = \DD[\eeta](\xx) + 
    \sum_{\ell=1}^{M} S[\llambda_\ell,\cc_\ell](\xx) +
    \sum_{\ell=1}^{M} R[\xi_\ell,\cc_\ell](\xx)\, , \quad \xx \in \Omega \, ,
\end{align}
where the density function, Stokeslets, and rotlets satisfy
\begin{subequations}
\label{eqn:completed_BIE}
\begin{align}
  \ff(\xx) &= -\frac{1}{2}\eeta(\xx) + \DD[\eeta](\xx) + 
      \sum_{\ell=1}^{M} S[\llambda_\ell,\cc_\ell](\xx) +
      \sum_{\ell=1}^{M} R[\xi_\ell,\cc_\ell](\xx) + \NN_0[\eeta](\xx)\, ,
      \quad \xx \in \bd\Omega \, , 
      \label{eqn:DLP}\\
  \llambda_\ell &= \frac{1}{2\pi}\int_{\gamma_\ell} \eeta(\yy) \,
  ds_\yy\, ,
  \quad \ell=1,\ldots,M \, ,
  \label{eqn:stokeslet} \\
  \xi_\ell &= \frac{1}{2\pi}\int_{\gamma_\ell} \yy^\perp \cdot \eeta(\yy)
  \, ds_\yy\, , \quad \ell=1,\ldots,M \, .
  \label{eqn:rotlet}
\end{align}
\end{subequations}
Here,
\begin{align*}
  \NN_0[\eeta](\xx) = \int_{\Gamma} 
    (\nn(\xx) \otimes \nn(\yy))\eeta(\yy) \, ds_\yy
\end{align*}
removes the rank one null space resulting from the flux-free requirement of $\ff$.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Computing the shear stress}
\label{sec:shearStressLP}
Once the density function, Stokeslets, and rotlets are computed, any desired quantity, including the shear stress, can be computed.  The shear stress is
\begin{align*}
  \tau = -\left(\nabla \uu + \nabla \uu^T \right)\nn \cdot \ss \, .
\end{align*}
The reader may notice an unusual sign convention above, but this is simply a consequence of our conventions for $\nn$ and $\ss$ (see Fig.~\ref{fig:schematic}).  The velocity field of the completed double-layer potential~\eqref{eqn:completed_DLP} includes velocities from three terms: the double-layer potential, the Stokeslets, and the rotlets.  We compute the deformation tensor $\ssigma = \frac{1}{2}(\nabla\uu + \nabla\uu^T)$ of these terms individually.

@@

To numerically compute the completed double-layer potential~\eqref{eqn:completed_BIE}, we discretize each of the $M$ bodies with  $N_\iin$ points, and the outer boundary with $N_\out$ points.  Then, $N = MN_\iin + N_\out$ is the total number of discretization points of $\bd\Omega$.  The double-layer potential in~\eqref{eqn:DLP} is discretized with the trapezoid rule as
\begin{align}
  \DD[\eeta](\xx_i) = \sum_{j=1}^{N} K(\xx_i,\xx_j) \eeta(\xx_j) 
      \Delta s_{j}\, , \quad i=1,\ldots,N \, ,
  \label{eqn:trapDLP}
\end{align}
where $\Delta s_j$ is the arclength term of $\bd\Omega$ at
$\xx_j$, and
\begin{align}
  K(\xx,\yy) = \frac{1}{\pi} \frac{\rr \cdot \nn}{\rho^2} 
      \frac{\rr \otimes \rr}{\rho^2} \, ,
\end{align}
where $\rr = \xx - \yy$ and $\rho = \|\rr\|$, is the kernel of the double-layer potential.  The diagonal term of~\eqref{eqn:trapDLP} is replaced with the limiting value of the double-layer potential
\begin{align}
  D(\xx,\xx) = -\frac{\kappa(\xx)}{2\pi}(\ss(\xx) \otimes \ss(\xx)) \, ,
\end{align}
where $\kappa(\xx)$ is the curvature and $\ss(\xx)$ is the unit tangent vector at $\xx \in \bd\Omega$.  The equations for the Stokeslets~\eqref{eqn:stokeslet} and the rotlets~\eqref{eqn:rotlet} are also discretized with the trapezoid rule.  Applying this discretization strategy to~\eqref{eqn:completed_BIE}, the dense linear system is solved is
\begin{equation}
  \label{eqn:completed_BIE_discretized}
  \begin{aligned}
    \ff(\xx_i) &= -\frac{1}{2}\ssigma(\xx_i) + \sum_{j=1}^{N} 
      K(\xx_i,\xx_j) \eeta(\xx_j) \Delta s_{j} + 
      \sum_{\ell=1}^M S[\llambda_\ell,\cc_\ell](\xx_i) \\
      &\hspace{53pt}
      +\sum_{\ell=1}^M R[\xi_\ell,\cc_\ell](\xx_i) 
      +\sum_{\xx_j \in \Gamma} (\nn(\xx_i) \otimes
      \nn(\xx_j))\eeta(\xx_j) \, , \\
    \llambda_\ell &= \sum_{\xx_j \in \gamma_\ell} \eeta(\xx_j) 
      \Delta s_j \, , \\ 
    \mu_\ell &= \sum_{\xx_j \in \gamma_\ell}
      {(\xx_j - \cc_\ell)}^\perp \cdot \eeta(\xx_j) \Delta s_j \, ,
  \end{aligned}
\end{equation}
for $i=1,\ldots,N$, $\ell=1,\ldots,M$.  We have applied the trapezoid rule since all the integrands are smooth and periodic, and this guarantees that the numerical solutions converges to the true solution with spectral accuracy~\cite{tre-wei2014}.

@@

For target points on the boundary of the geometry, $\xx \in \bd \Omega$, there is an additional term resulting from a jump in the pressure double-layer potential~\cite{poz1992}
\begin{align}
  p(\xx)= \pderiv{\eeta}{\ss} \cdot \ss + p^{\DD}(\xx) + 
              p^{S}(\xx)\, , \quad \xx \in \bd\Omega \, .
  \label{eqn:pressureJump}
\end{align}
Similar to the shear stress, the singularity of the pressure corresponding to the double-layer potential~\eqref{eqn:pressureDLP} is too strong when $\xx \in \bd\Omega$ for the trapezoid rule to be directly applied.  Therefore, we again subtract a constant density function and write the pressure as
\begin{equation}
\begin{aligned}
  p^\DD(\xx) = -\frac{1}{\pi}\int_{\bd\Omega} \frac{1}{\rho^2}
    \left(I - 2 \frac{\rr \otimes \rr}{\rho^2}\right) 
    \nn \cdot (\eeta(\yy) - \eeta(\xx)) \,ds_\yy\, , 
    \quad \xx \in \bd\Omega\, .
\end{aligned}
\end{equation}
Then, identical to how the shear stress is computed~\eqref{eqn:stressOddEven}, odd-even integration is used to evaluate this layer potential for pressure, and the jump condition in~\eqref{eqn:pressureJump} is computed with Fourier differentiation.

@@

As a longer term goal, we would like to include the effect of sedimentation to more faithfully capture the physics of a porous medium. Sedimentation requires overcoming a few challenges. First, the bodies must be mobile, as has been accomplished in previous work for dense concentrations of deformable and rigid bodies~\cite{qua-bir2014a, cor-gre-rac-vee2017, cor-vee2017, kli-tor2016b, rac-gre2016}. Second, contacts and overlap between bodies must be prevented. This second requirement has been recognized as quite a challenge, with many novel methods developed to treat it~\cite{lu-rah-zor2017, kab-qua-bir2018, san-mo1994}. Incorporation of one or more of these schemes, will enable simulations of a porous medium undergoing simultaneous erosion and sedimentation.

>>>>

Recent work in the high-Reynolds-number regime has highlighted the importance of a shape-flow feedback that occurs during erosion~\cite{ris-moo-chi-she-zha2012, moo-ris-chi-zha-she2013, hewett2017evolution, moore2017riemann} and the related processes of dissolution and melting~\cite{Huang2015, kondratiuk2015steadily, rycroft2016asymmetric, cohen2016erosion, hewett2017pear, claudin2017dissolution}.  In these cases, the flow alters the morphology of the immersed structures, which in turn modifies the surrounding flow, and so on. Such processes of `sculpting with flow'~\cite{ristroph2018sculpting} have been examined in the context of a single body~\cite{ris-moo-chi-she-zha2012, moore2017riemann} and, more recently, multiple bodies whose shape evolution is coupled through the flow~\cite{hewett2017evolution}. From this viewpoint, the example of an eroding porous medium presents new opportunities and challenges. First, since the medium is composed of many individual constituents, multi-body interactions must be considered. Second, the long-range effects inherent to Stokes flow implies that the shape-evolution of bodies are even more strongly coupled than in any other regime.
% in particular close-contacts?

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\paragraph{Contributions}
To perform this simulation of Stokesian erosion of a porous media, two fast and accurate methods are employed---one to solve the fluid equations and the other to time advance the eroding grains.  The fluid equations are reformulated in terms of a boundary integral equation (BIE)~\cite{poz1992}.  Since the boundary conditions are Dirichlet, we use a double-layer potential formulation so that the resulting BIE is second-kind.  This second-kind BIE is discretized with the spectrally accurate trapezoid rule~\cite{tre-wei2014} which results in a well-conditioned dense linear system for an unknown density function, and the linear system is solved using fast-multipole-accelerated GMRES~\cite{saa-sch1986}.  Once the density function is computed, the shear stress is computed with a spectrally accurate odd-even quadrature rule~\cite{sid-isr1988}.  In addition, the pressure and vorticity are computed in the fluid bulk (Fig.~\ref{fig:50bodies}) with an upsampled trapezoid rule to assist with visualizing the effect of erosion in porous media.

@@

\paragraph{Related Work}
In closely related work, Mitchell \& Spagnolie used a traction integral equation formulation to simulate viscous erosion of a single body in three dimensions~\cite{mit-spa2016}. The flexibility of their formulation enabled simulations of a {\em mobile} body eroding in an {\em aribtrary} background flow (e.g.~uniform flow, shear flow, etc.).  Another advantage of the traction formulation is that is directly provides the shear stress that is needed to evolve surfaces.  In the present work, we aim to simulate the erosion of a porous medium and, further, choose to fix the bodies to their initial locations. Thus, rather than a {\em single, mobile} body as considered in~\cite{mit-spa2016}, we aim to simulate the erosion of {\em many, immobile} bodies. Although the traction formulation has several attractive features, we instead employ a standard double-layer formulation since it allows us to easily calculate bulk quantities (e.g.~vorticity, pressure, and velocity) that will assist our understanding of the erosion process.

@@

Since the double-layer potential is not able to represent force- and torque-free velocity fields,  we complete the double-layer potential with Stokeslets and rotlets~\cite{pow-mir1987, pow1993}
\begin{align}
  S[\llambda_\ell,\cc_\ell] = \frac{1}{4\pi} \left(-\log \rho \II + 
    \frac{\rr \otimes \rr}{\rho^2} \right)\llambda_\ell
  \quad \text{and} \quad
  R[\xi_\ell,\cc_\ell] = \frac{\rr^\perp}{\rho^2}\xi_\ell\, , 
  \quad \ell=1,\ldots,M \, ,
  \label{eqn:stokeslet_rotlet}
\end{align}
where  $\rr = \xx - \cc_\ell$, $\cc_\ell$ is a point inside the $\ell^{th}$ grain, $\rr^\perp = (r_2,-r_1)$, and $\rho = \|\rr\|$.  Then, the solution of~\eqref{eqn:stokes} is
\begin{align}
  \label{eqn:completed_DLP}
  \uu(\xx) = \DD[\eeta](\xx) + 
    \sum_{\ell=1}^{M} S[\llambda_\ell,\cc_\ell](\xx) +
    \sum_{\ell=1}^{M} R[\xi_\ell,\cc_\ell](\xx)\, , \quad \xx \in \Omega \, ,
\end{align}
where the density function, Stokeslets, and rotlets satisfy
\begin{subequations}
\label{eqn:completed_BIE}
\begin{align}
  \ff(\xx) &= -\frac{1}{2}\eeta(\xx) + \DD[\eeta](\xx) + 
      \sum_{\ell=1}^{M} S[\llambda_\ell,\cc_\ell](\xx) +
      \sum_{\ell=1}^{M} R[\xi_\ell,\cc_\ell](\xx) + \NN_0[\eeta](\xx)\, ,
      \quad \xx \in \bd\Omega \, , 
      \label{eqn:DLP}\\
  \llambda_\ell &= \frac{1}{2\pi}\int_{\gamma_\ell} \eeta(\yy) \,
  ds_\yy\, ,
  \quad \ell=1,\ldots,M \, ,
  \label{eqn:stokeslet} \\
  \xi_\ell &= \frac{1}{2\pi}\int_{\gamma_\ell} \yy^\perp \cdot \eeta(\yy)
  \, ds_\yy\, , \quad \ell=1,\ldots,M \, .
  \label{eqn:rotlet}
\end{align}
\end{subequations}
Here,
\begin{align}
  \NN_0[\eeta](\xx) = \int_{\Gamma} 
    (\nn(\xx) \otimes \nn(\yy))\eeta(\yy) \, ds_\yy
\end{align}
removes the rank one null space resulting from the flux-free requirement of $\ff$.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Computing the shear stress}
\label{sec:shearStressLP}
Once the density function, Stokeslets, and rotlets are computed, any desired quantity, including the shear stress, can be computed.  The shear stress is
\begin{align}
  \tau = -\left(\nabla \uu + \nabla \uu^T \right)\nn \cdot \ss \, .
\end{align}
The reader may notice an unusual sign convention above, but this is simply a consequence of our conventions for $\nn$ and $\ss$ (see Fig.~\ref{fig:schematic}).  The velocity field of the completed double-layer potential~\eqref{eqn:completed_DLP} includes velocities from three terms: the double-layer potential, the Stokeslets, and the rotlets.  We compute the deformation tensor $\ssigma = \frac{1}{2}(\nabla\uu + \nabla\uu^T)$ of these terms individually.

@@

To numerically compute the completed double-layer potential~\eqref{eqn:completed_BIE}, we discretize each of the $M$ bodies with  $N_\iin$ points, and the outer boundary with $N_\out$ points.  Then, $N = MN_\iin + N_\out$ is the total number of discretization points of $\bd\Omega$.  The double-layer potential in~\eqref{eqn:DLP} is discretized with the trapezoid rule as
\begin{align}
  \DD[\eeta](\xx_i) = \sum_{j=1}^{N} K(\xx_i,\xx_j) \eeta(\xx_j) 
      \Delta s_{j}\, , \quad i=1,\ldots,N \, ,
  \label{eqn:trapDLP}
\end{align}
where $\Delta s_j$ is the arclength term of $\bd\Omega$ at
$\xx_j$, and
\begin{align}
  K(\xx,\yy) = \frac{1}{\pi} \frac{\rr \cdot \nn}{\rho^2} 
      \frac{\rr \otimes \rr}{\rho^2} \, ,
\end{align}
where $\rr = \xx - \yy$ and $\rho = \|\rr\|$, is the kernel of the double-layer potential.  The diagonal term of~\eqref{eqn:trapDLP} is replaced with the limiting value of the double-layer potential
\begin{align}
  D(\xx,\xx) = -\frac{\kappa(\xx)}{2\pi}(\ss(\xx) \otimes \ss(\xx)) \, ,
\end{align}
where $\kappa(\xx)$ is the curvature and $\ss(\xx)$ is the unit tangent vector at $\xx \in \bd\Omega$.  The equations for the Stokeslets~\eqref{eqn:stokeslet} and the rotlets~\eqref{eqn:rotlet} are also discretized with the trapezoid rule.  Applying this discretization strategy to~\eqref{eqn:completed_BIE}, the dense linear system that solved is
\begin{equation}
  \label{eqn:completed_BIE_discretized}
  \begin{aligned}
    \ff(\xx_i) &= -\frac{1}{2}\ssigma(\xx_i) + \sum_{j=1}^{N} 
      K(\xx_i,\xx_j) \eeta(\xx_j) \Delta s_{j} + 
      \sum_{\ell=1}^M S[\llambda_\ell,\cc_\ell](\xx_i) \\
      &\hspace{53pt}
      +\sum_{\ell=1}^M R[\xi_\ell,\cc_\ell](\xx_i) 
      +\sum_{\xx_j \in \Gamma} (\nn(\xx_i) \otimes
      \nn(\xx_j))\eeta(\xx_j) \, , \\
    \llambda_\ell &= \sum_{\xx_j \in \gamma_\ell} \eeta(\xx_j) 
      \Delta s_j \, , \\ 
    \xi_\ell &= \sum_{\xx_j \in \gamma_\ell}
      {(\xx_j - \cc_\ell)}^\perp \cdot \eeta(\xx_j) \Delta s_j \, ,
  \end{aligned}
\end{equation}
for $i=1,\ldots,N$, $\ell=1,\ldots,M$.  We have applied the trapezoid rule since all the integrands are smooth and periodic, and this guarantees that the numerical solutions converges to the true solution with spectral accuracy~\cite{tre-wei2014}.

@@

For target points on the boundary of the geometry, $\xx \in \bd \Omega$, there is an additional term resulting from a jump in the pressure double-layer potential~\cite{poz1992}
\begin{align}
  p(\xx)= \pderiv{\eeta}{\ss} \cdot \ss + p^{\DD}(\xx) + 
              p^{S}(\xx)\, , \quad \xx \in \bd\Omega \, .
  \label{eqn:pressureJump}
\end{align}
Similar to the shear stress, the singularity of the pressure corresponding to the double-layer potential~\eqref{eqn:pressureDLP} is too strong when $\xx \in \bd\Omega$ for the trapezoid rule to be directly applied.  Therefore, we again subtract a constant density function and write the pressure as
\begin{equation}
\begin{aligned}
  p^\DD(\xx) = -\frac{1}{\pi}\int_{\bd\Omega} \frac{1}{\rho^2}
    \left(I - 2 \frac{\rr \otimes \rr}{\rho^2}\right) 
    \nn \cdot (\eeta(\yy) - \eeta(\xx)) \,ds_\yy\, , 
    \quad \xx \in \bd\Omega\, .
\end{aligned}
\end{equation}
Then, identical to how the shear stress is computed in~\eqref{eqn:stressOddEven}, odd-even integration is used to evaluate this layer potential for pressure, and the jump condition in~\eqref{eqn:pressureJump} is computed with Fourier differentiation.

@@

As a longer term goal, we would like to include the effect of sedimentation to more faithfully capture the physics of a porous medium.  Sedimentation requires overcoming a few challenges. First, the bodies must be mobile, as has been accomplished in previous work for dense concentrations of deformable and rigid bodies~\cite{qua-bir2014a, cor-gre-rac-vee2017, cor-vee2017, kli-tor2016b, rac-gre2016, qua-bir2016}. Second, contacts and overlap between bodies must be prevented. This second requirement has been recognized as quite a challenge, with many novel methods developed to treat it~\cite{lu-rah-zor2017, kab-qua-bir2018, san-mo1994}. Incorporation of one or more of these schemes, will enable simulations of a porous medium undergoing simultaneous erosion and sedimentation.
